---
title: "... Exercise 1 - Independent Exercise - Customize a PCA plot"
author: "UM Bioinformatics Core"
date: "`r Sys.Date()`"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            toc_depth: 4
            toc_float: true
            number_sections: false
            fig_caption: true
            markdown: GFM
            code_download: true
---

<style type="text/css">
body, td {
   font-size: 18px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}
</style>

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09a-")
```

```{r Modules, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(DESeq2)
library(ggplot2)
library(tidyr)
library(dplyr)
library(matrixStats)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
# load("rdata/RunningData.RData")
```

Estimated time: **15 Minutes**


# Motivation

Customizing plots can help us see patterns in the data or make the claim(s) based on the data represented clearer.

# Instructions

- Work independently in the main room, posting any questions that arise to slack.
- Recommendations for writing your own code:
  - Read function documentation
  - Test out ideas - it's okay to make mistakes and generate errors
  - Use a search engine to look up errors or recommended solutions using keywords
- We'll review possible solutions after time is up as a group.

# Exercise

Try doing the following to the `pca_plot`, starting with the "most popular" request and moving on to other customizations if you have time:

> * Add a title and subtitle to the plot
> * Update the color palette to be color-blind friendly
> * Add labels to show which samples correspond to which points
> * Use shape instead of color to indicate groups on the PCA plot.
> * _Challenge_: Change the legend title to "Iron Status". 

# Example

Here is a copy of the code we just tested together to 1) pull the underlying data from the PCA function and 2) change the theme of our PCA plot to black and white.

```{r test_PCA}
pcaData <- plotPCA(rld, intgroup=c("condition"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar")) # store PC axes (% variance)

# create custom plot object
PCACustom <- ggplot(pcaData, aes(PC1, PC2, color=condition)) +
  geom_point(size=3) +
  coord_fixed() +
  theme_bw()

# add percentVar labels to *displayed plot*
PCACustom + 
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance"))

# add percentVar labels to *stored plot object*
PCACustom2 <- PCACustom + 
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance"))
```

# Steps

* Add a title and subtitle to the ggplot plot 
  * Hint: use the `labs()` function or search for [examples](https://www.datanovia.com/en/blog/ggplot-title-subtitle-and-caption/)

* Add labels to show which samples correspond to which points
  * Hint: use the 
  * See [example in this support post](https://support.bioconductor.org/p/90791/) and/or [apply more general recommendations from this R graph gallery](https://r-graph-gallery.com/275-add-text-labels-with-ggplot2.html)
  
* Make our color palette more color-blind friendly (with `RColorBrewer`)
  * When considering color choices, be aware of the [prevalence and types of color-blindness](https://medium.com/version-1/simulating-visualisations-in-r-for-colour-blindness-2faddde63695) and how to choose more [easily distinguishable colors](http://bconnelly.net/posts/creating_colorblind-friendly_figures/) and check the final figure for sufficient contrasts with a [simulator tool](https://www.color-blindness.com/coblis-color-blindness-simulator/). 
  * There are pre-designed [color-blind friendly palettes available](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette)
  
* Use shape instead of color to indicate groups on the PCA plot. 
  * Hint: since color and shape are both mapping aspects, so we'll need to re-specify shape instead of color instead of adding an additional layers like the other prompts 
  * [These R graph gallery examples](https://r-graph-gallery.com/274-map-a-variable-to-ggplot2-scatterplot.html) could be a good place to start

* _Challenge_: Change the legend title to "Iron Status". 
  * Hint, you can do this with the `labs()` function too, using the corresponding aesthetic mapping (e.g. "color").


# Saving the result

If time permits, discuss with your group how you might save this heatmap. Hint: Look at the parameters for the function in `?pheatmap`. Alternatively, consider how we saved the PCA in the previous module.

```{r WriteOut.RData, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#Hidden code block to write out data for knitting
# save.image(file = "rdata/RunningData.RData")
```
